---
title: "Lab6"
author: "Charlie Curtin"
date: "2023-03-01"
output: html_document
---

## Case Study: Eel Distribution Modeling

This week's lab follows a project modeling the eel species Anguilla australis described by Elith et al. (2008). There are two data sets for this lab.  You'll use one for training and evaluating your model, and you'll use your model to make predictions predictions on the other.  Then you'll compare your model's performance to the model used by Elith et al.

## Data

Grab the training data sets (eel.model.data.csv, eel.eval.data.csv) from github here:
https://github.com/MaRo406/eds-232-machine-learning/blob/main/data 

```{r, message = FALSE, warning = FALSE}
# load packages
library(tidyverse)
library(here)
library(parsnip)
library(caret)
library(tidymodels)
```


```{r, message = FALSE, warning = FALSE}
# read in the data
eel_model <- read_csv(here("labs", "data", "eel.model.data.csv")) %>% 
  mutate(Angaus = as.factor(Angaus))

eel_eval <- read_csv(here("labs", "data", "eel.eval.data.csv"))
```


### Split and Resample

Split the model data (eel.model.data.csv) into a training and test set, stratified by outcome score (Angaus). Use 10-fold CV to resample the training set.

```{r}
set.seed(345)

# split the model data
eel_split <- initial_split(eel_model, strata = Angaus)
eel_train <- training(eel_split)
eel_test <- testing(eel_split)

# specify k-folds resampling with 10 folds
cv_folds <- vfold_cv(eel_train, v = 10)
```


### Preprocess

Create a recipe to prepare your data for the XGBoost model

```{r}
# specify a recipe
eel_rec <- recipe(Angaus ~ ., data = eel_train) %>% 
  # turn nominal variables into dummy variables
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>% 
  step_normalize(all_numeric_predictors())
  
```


## Tuning XGBoost

### Tune Learning Rate

Following the XGBoost tuning strategy outlined in lecture, first we conduct tuning on just the learning rate parameter:

1.  Create a model specification using {xgboost} for the estimation

```{r}
# specify an xgboost model with learn_rate set to tune
xgb_spec <- boost_tree(learn_rate = tune()) %>% 
  set_mode("classification") %>% 
  set_engine("xgboost")

# create a workflow with our model
xgb_wf <- workflow() %>% 
  add_model(xgb_spec) %>% 
  add_recipe(eel_rec)
```


2.  Set up a grid to tune your model by using a range of learning rate parameter values: expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))

```{r}
# set up grid to tune our model
system.time(
  xgb_tune <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid = expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))
))

# assess accuracy
```


-   Use appropriate metrics argument(s) - Computational efficiency becomes a factor as models get more complex and data get larger. Record the time it takes to run. Do this for each tuning phase you run.You could use {tictoc} or Sys.time().

3.  Show the performance of the best models and the estimates for the learning rate parameter values associated with each.

### Tune Tree Parameters

1.  Create a new specification where you set the learning rate (which you already optimized) and tune the tree parameters.

2.  Set up a tuning grid. This time use grid_latin_hypercube() to get a representative sampling of the parameter space

3.  Show the performance of the best models and the estimates for the tree parameter values associated with each.

```{r}
tune_grid <- grid_latin_hypercube()
```


### Tune Stochastic Parameters

1.  Create a new specification where you set the learning rate and tree parameters (which you already optimized) and tune the stochastic parameters.

2.  Set up a tuning grid. Use grid_latin_hypercube() again.

3.  Show the performance of the best models and the estimates for the tree parameter values associated with each.

## Finalize workflow and make final prediction

1.  How well did your model perform? What types of errors did it make?

## Fit your model the evaluation data and compare performance

1.  Now used your final model to predict on the other dataset (eval.data.csv)

2.  How does your model perform on this data?

3.  How do your results compare to those of Elith et al.?

-   Use {vip} to compare variable importance
-   What do your variable importance results tell you about the distribution of this eel species?
